---
title: "ENV 316: R Markdown Tutorial "
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Howdy!

We understand that virtual classes may not have been what you had in mind when you signed up for a course titled *Laboratory and Field Methods in Environmental Science*, but it does provide us with a great oppertunity to really dive into what seperates science from a walk in the woods: writting stuff down! As you've probably already heard, the ENV 316 course will make prodigious use of R and RStudio as we explore concepts of environmental chemistry and ecology. Since you're already using R and RStudio for your data analysis and manipulations, we're encouraging you to submit your work in R Markdown. 

The aim of this document is to briefly explain what R Markdown is, why you should use it (hint: it'll make everyone's lives easier), and how to create simple documents for this course. 

## First off, what is R Markdown? 

In a nutshell, R Markdown allows you to analyse your data with R and write your report in the same place (this document writen with R Markdown). This has loads of benefits including increased reproducibility, and streamlined thinking. No more flipping back and forth between code and writing to figure out what's going on.  For example, 

```{r}
# Look at me go mom
x <- 2+2
x
```


